\name: A hierarchical Bayesian model for Supernova Cosmology

### 2018-07-28 papers/methods/paper.tex
More updates from feedback with Dan

At the most fundamental level, a supernova analysis is simply a mapping from an underlying population onto an observed population, where cosmology is encoded directly in the mapping function. The difficulty arises both in adequately describing the biases in the mapping function, and in adding sufficient, physically motivated flexibility in both of these populations whilst not adding \textit{too} much flexibility, such that model fitting becomes pathological due to increasing parameter degeneracies within the model. In the following sections, we will describe these layers, mapping functions, and occurrences of these fatal pathologies.

@@

Like most of the BHM methods introduced previously, we work from the summary statistics, where each observed supernova has an apparent magnitude $\hat{m}_B$, stretch $\hat{x}_1$ and colour $\hat{c}$, with uncertainty $\cov$ on those values. Additionally, each supernova has an observed redshift $\hat{z}$ and a host galaxy mass associated with it, $\hat{m}$, where the mass measurement takes the form of a probability of being above $10^{10}$ solar masses. We will also have a probability of being a Type Ia, $\hat{p}$. Our set of observables is therefore given as $\lbrace \hat{m}_B, \hat{x}_1, \hat{c}, \hat{z}, \hat{p}, \hat{m}, \cov \rbrace$, as shown in the probabilistic graphical model (PGM) in Figure \ref{fig:pgm}.

As we are focused on the spectroscopically confirmed supernovae for this iteration of the method, we assume the observed redshift $\hat{z}$ is the true redshift $z$ such that $P(\hat{z}|z) = \delta(\hat{z} - z)$. Potential sources of redshift error (such as peculiar velocities) is taken into account not via uncertainty on redshift (which is technically challenging to implement) but instead uncertainty on distance modulus. This is discussed further in Section \ref{sec:systreat}. Similarly, we take the mass probability estimate $\hat{m}$ as correct, and do not model a latent variable. One of the strengths of this model (and the {\rubin} analysis) is that for future data sets where supernovae have been classified photometrically, and we expect some misclassification and misidentification of the host galaxies, those can naturally be modelled and taken into account.

@@

To help confuse things, we often consider models that are parametrised as $\theta = \lbrace m_B, x_1, c \rbrace = \eta$, so that the probability of derived data is expressed as

@@

We model the extra dispersion only in colour, and do so by adding extra independent uncertainty on the colour observation. Fully covariant extra dispersion on $\lbrace m_B, x_1, c \rbrace$ was also tested, but it showed negligible improvement over just colour dispersion and was far more computationally inefficient. As shown in \citep{Kessler2013}, the extra colour dispersion shows heavy redshift dependence. As the extra dispersion may arise from underlying supernova physics, rather than Malmquist bias, we decide to incorporate redshift dependence in our extra uncertainty. We thus add $\kappa_0 + \kappa_1 z$ to our observed colour uncertainty (in quadrature), where $\kappa_0$ and $\kappa_1$ are subject to strong Cauchy priors centered on zero and with width $0.002$, to enforce minimum possible extra uncertainty in our model. The $\kappa$ parameters are tightly constrained as they are highly degenerate with the width of the intrinsic colour population $\sigma_c$, and the specific value of this prior was determined from running many iterations of simulation fitting with differing prior strengths on our simple data set with simulated extra dispersion according to the results of \citet{Kessler2013}. The value $0.002$ can be varied by a factor of approximately three in either direction whilst giving stable results, with larger values (more than a factor of three) for the prior resulted in incorrect determination of $\sigma_c$, whilst smaller values resulted in being unable to capture the extra dispersion  Each survey has its own set of $\kappa$ parameters. As such, our combined covariance on the observation $\hat{\eta}$ is given by $\mathcal{C}_{\rm tot} = \mathcal{C} + {\rm DiagMatrix}\left[0, 0, (\kappa_0 + \kappa_1 z)^2\right]$.

@@

The selection functions are fit to apparent magnitude efficiency ratios calculated from SNANA simulations, by calculating an efficiency ratio as a function of apparent magnitude. Uncertainty of the Malmquist bias from the imperfection of the analytic function is incorporated into the fit. Uncertainty was uniformly added in quadrature (similar to adding intrinsic dispersion as a fit parameter) to the efficiency ratio data from our simulations until the reduced $\chi^2$ of the analytic fit reached one, allowing us to extract an uncertainty covariance matrix for our analytic fits to either the error function or the skew normal. We can thus model a parametrised selection effect by including deviations from the determined mean selection function parameters with parameter vector $\Delta S$, and apply a normal prior on this parameter as per the determined uncertainty covariance matrix.

>>>>

At the most fundamental level, a supernova cosmology analysis is simply a mapping from an underlying population onto an observed population, where cosmology is encoded directly in the mapping function. The difficulty arises both in adequately describing the biases in the mapping function, and in adding sufficient, physically motivated flexibility in both of these populations whilst not adding \textit{too} much flexibility, such that model fitting becomes pathological due to increasing parameter degeneracies within the model. In the following sections, we will describe these layers, mapping functions, and occurrences of these fatal pathologies.

@@

Like most of the BHM methods introduced previously, we work from the summary statistics, where each observed supernova has a flux measurement $\hat{m}_B$ (which is analogous to apparent magnitude), stretch $\hat{x}_1$ and colour $\hat{c}$, with uncertainty $\cov$ on those values. Additionally, each supernova has an observed redshift $\hat{z}$ and a host galaxy mass associated with it, $\hat{m}$, where the mass measurement takes the form of a probability of being above $10^{10}$ solar masses. We will also have a probability of each supernovae being a Type Ia, $\hat{p}$. Our set of observables is therefore given as $\lbrace \hat{m}_B, \hat{x}_1, \hat{c}, \hat{z}, \hat{p}, \hat{m}, \cov \rbrace$, as shown in the probabilistic graphical model (PGM) in Figure \ref{fig:pgm}.

As we are focused on the spectroscopically confirmed supernovae for this iteration of the method, we assume the observed redshift $\hat{z}$ is the true redshift $z$ such that $P(\hat{z}|z) = \delta(\hat{z} - z)$. Potential sources of redshift error (such as peculiar velocities) are taken into account not via uncertainty on redshift (which is technically challenging to implement) but instead uncertainty on distance modulus. This is discussed further in Section \ref{sec:systreat}. Similarly, we take the mass probability estimate $\hat{m}$ as correct, and do not model a latent variable to represent uncertainty on the probability estimate. One of the strengths of this model (and the {\rubin} analysis) is that for future data sets where supernovae have been classified photometrically, and we expect some misclassification and misidentification of the host galaxies, those can naturally be modelled and taken into account.

@@

We often consider models that are parametrised as $\theta = \lbrace m_B, x_1, c \rbrace = \eta$, so that the probability of derived data is expressed as

@@

We model the extra dispersion only in colour, and do so by adding extra independent uncertainty on the colour observation. Fully covariant extra dispersion on $\lbrace m_B, x_1, c \rbrace$ was also tested, but it showed negligible improvement in recovering unbiased cosmology over just colour dispersion, and was far more computationally inefficient. We note here that we do model dispersion in magnitude, but this is done at the level of underlying populations, not observed populations. As shown in \citep{Kessler2013}, the extra colour dispersion shows heavy redshift dependence. As the extra dispersion may arise from underlying supernova physics, rather than Malmquist bias, we decide to incorporate redshift dependence in our extra uncertainty. We thus add $\kappa_0 + \kappa_1 z$ to our observed colour uncertainty (in quadrature), where $\kappa_0$ and $\kappa_1$ are subject to strong Cauchy priors centered on zero and with width $0.002$, to enforce minimum possible extra uncertainty in our model. The $\kappa$ parameters are tightly constrained as they are highly degenerate with the width of the intrinsic colour population $\sigma_c$, and the specific value of this prior was determined from running many iterations of simulation fitting with differing prior strengths on our simple data set with simulated extra dispersion according to the results of \citet{Kessler2013}. The value $0.002$ can be varied by a factor of approximately three in either direction whilst giving stable results, with larger values (more than a factor of three) for the prior resulted in incorrect determination of $\sigma_c$, whilst smaller values resulted in being unable to capture the extra dispersion  Each survey has its own set of $\kappa$ parameters. As such, our combined covariance on the observation $\hat{\eta}$ is given by $\mathcal{C}_{\rm tot} = \mathcal{C} + {\rm DiagMatrix}\left[0, 0, (\kappa_0 + \kappa_1 z)^2\right]$.

@@

The selection functions are fit to apparent magnitude efficiency ratios calculated from SNANA simulations, by calculating an efficiency ratio as a function of apparent magnitude. Uncertainty of the Malmquist bias from the imperfection of the analytic function is incorporated into the fit. Uncertainty was uniformly added in quadrature (similar to adding intrinsic dispersion as a fit parameter) to the efficiency ratio data from our simulations until the reduced $\chi^2$ of the analytic fit reached one, allowing us to extract an uncertainty covariance matrix for our analytic fits to either the error function or the skew normal. We can thus model a parametrised selection effect by including deviations from the determined mean selection function parameters with parameter vector $\Delta S$, and apply a normal prior on this parameter as per the determined uncertainty covariance matrix. {\green Whilst this uncertainty encapsulates the potential error from the simulations not matching the analytic approximations, it does not cover potential variations of the selection function at the top level - varying cosmology or spectroscopic efficiency.}
